---
title: "Introduction to proteoDeconv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to proteoDeconv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignette.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette describes a basic workflow that can be used to run the proteoDeconv package.

## Loading required packages

First, let's load the necessary packages:

```{r setup, message=FALSE}
library(proteoDeconv)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Example data overview

`proteoDeconv` includes two example datasets, which are subsets of our dataset PXD056050:

1. Pure cell type samples: Proteomic measurements from isolated CD8+ T cells and monocytes.
2. Mixed samples: Proteomic data from experimental mixtures containing 50% CD8+ T cells and 50% monocytes.

Let's load these datasets:

```{r}
pure_data_file <- system.file("extdata", "pure_samples_matrix.rds", package = "proteoDeconv")
pure_samples <- readRDS(pure_data_file)

mix_data_file <- system.file("extdata", "mixed_samples_matrix.rds", package = "proteoDeconv")
mixed_samples <- readRDS(mix_data_file)
```

Let's examine the structure of the pure samples data:

```{r}
dim(pure_samples)

head(pure_samples)
```

The data matrix contains protein measurements (rows) across different samples (columns). The sample names indicate the cell type (CD8 for CD8+ T cells and Mono for monocytes).

## Data preprocessing

Proper preprocessing is important for accurate deconvolution. The preprocessing pipeline in `proteoDeconv` includes several important steps:

```{r}
pure_samples_preprocessed <- pure_samples |>
  # Extract first gene group when multiple gene groups are present
  extract_identifiers() |>
  # Update gene symbols to current HGNC nomenclature
  update_gene_symbols() |>
  # Handle missing values by imputing with the lowest observed value
  handle_missing_values() |>
  # Resolve duplicate genes by keeping the row with highest median value
  handle_duplicates() |>
  # Scale the data (TPM-like normalization)
  convert_to_tpm()

# Apply the same preprocessing to mixed samples
mixed_samples_preprocessed <- mixed_samples |>
  extract_identifiers() |>
  update_gene_symbols() |>
  handle_missing_values() |>
  handle_duplicates() |>
  convert_to_tpm()
```

Let's examine the preprocessed data:

```{r}
head(pure_samples_preprocessed)
```

### Preprocessing options

`proteoDeconv` offers several options for each preprocessing step, which have been evaluated for their impact on deconvolution performance. For further information on the options, please refer to the function reference.

## Creating a signature matrix

A signature matrix contains cell type-specific marker proteins that will be used for the deconvolution. Signature matrices represent the reference profiles of pure cell populations.

To create one, we first need to establish which samples correspond to which cell types:

```{r}
mapping_rules <- list(
  "CD8+ T cells" = "CD8", # Samples containing "CD8" will be classified as CD8+ T cells
  "Monocytes" = "Mono" # Samples containing "Mono" will be classified as Monocytes
)

phenoclasses <- create_phenoclasses(
  pure_samples_preprocessed,
  mapping_rules,
  verbose = TRUE
)

head(phenoclasses)
```

The phenoclasses matrix shows which cell type each sample represents, which is required for the signature matrix generation.

Now, we can create the signature matrix using CIBERSORTx:

```{r eval=FALSE}
# Note: This requires a CIBERSORTx token to be set in .Renviron
signature_matrix <- create_signature_matrix(
  refsample = pure_samples_preprocessed,
  phenoclasses = phenoclasses,
  g_min = 200,
  g_max = 400,
  q_value = 0.01
)
```

```{r echo=FALSE}
signature_file <- system.file("extdata", "cd8t_mono_signature_matrix.rds", package = "proteoDeconv")
signature_matrix <- readRDS(signature_file)
```

Let's examine the signature matrix:

```{r}
dim(signature_matrix)

# Look at the first rows
head(signature_matrix)
```

The signature matrix contains proteins (rows) with their expression values in each cell type (columns). These proteins were selected based on their ability to differentiate between CD8+ T cells and monocytes.

## Performing deconvolution

Now we can deconvolute our mixed samples to estimate their cell type composition. The `proteoDeconv` package supports multiple algorithms that have been evaluated for their performance with proteomics data. We'll use the EPIC algorithm here [@racleSimultaneousEnumerationCancer2017]:

```{r}
results <- deconvolute(
  algorithm = "epic",
  data = mixed_samples_preprocessed,
  signature = signature_matrix,
  with_other_cells = FALSE # No need to calculate other cells in this case
)

print(results)
```

The results show the estimated proportion of each cell type in each mixed sample.

### Available deconvolution algorithms

`proteoDeconv` supports multiple deconvolution algorithms:

- **EPIC** [@racleSimultaneousEnumerationCancer2017]: A reference-based method that uses constrained least-squares regression
- **CIBERSORT** [@newmanRobustEnumerationCell2015]: One of the first widely-used deconvolution tools, uses support vector regression (requires downloading the CIBERSORT.R script from the CIBERSORT website)
- **CIBERSORTx** [@newmanDeterminingCellType2019]: An enhanced version of CIBERSORT that includes batch correction features (requires a token and Docker)
- **BayesDeBulk** [@petraliaBayesDeBulkFlexibleBayesian2023]: A Bayesian approach specifically designed for proteomics deconvolution

To use a different algorithm, simply change the `algorithm` parameter:

```{r, eval=FALSE}
# Example using a different algorithm
results_cibersort <- deconvolute(
  algorithm = "cibersort",
  data = mixed_samples_preprocessed,
  signature = signature_matrix
)
```

## Formatting and visualizing results

For easier visualization, let's convert the results to a tidy format:

```{r}
results_tidy <- as_tibble(results, rownames = "sample") |>
  pivot_longer(
    cols = -sample,
    names_to = "cell_type",
    values_to = "cell_fraction"
  )

head(results_tidy)
```

Now, let's create a bar chart to visualize the cell type proportions:

```{r fig.height=4}
ggplot(results_tidy, aes(x = sample, y = cell_fraction, fill = cell_type)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Sample",
    y = "Estimated Proportion",
    fill = "Cell Type"
  )
```

## Conclusion

This vignette demonstrated the basic workflow for proteomics deconvolution using the `proteoDeconv` package:

1. Load and preprocess your data
2. Create a cell type signature matrix
3. Perform deconvolution on mixed samples
4. Visualize and interpret the results

For more advanced usage, refer to the package documentation and function reference.

## References
